\documentclass[acmsmall,screen,review,anonymous]{acmart}

\setcopyright{none}
\acmConference[POPL '27]{ACM SIGPLAN Symposium on Principles of Programming Languages}{January 2027}{}
\acmYear{2027}
\acmDOI{}
\acmISBN{}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{xspace}

\newcommand{\system}{CiPR-FTQC\xspace}
\newcommand{\lang}{FTQP\xspace}
\newcommand{\judgment}{\mathcal{K};\Gamma;\Delta \vdash t : \Gamma';\Delta' \;!\; E}
\newcommand{\code}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\gate}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\res}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\mode}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\cert}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\Cap}{\ensuremath{\mathsf{Cap}}}
\newcommand{\Switch}{\ensuremath{\mathsf{Switch}}}
\newcommand{\Factory}{\ensuremath{\mathsf{Factory}}}
\newcommand{\Backend}{\ensuremath{\mathsf{Backend}}}
\newcommand{\Checked}{\cert{Checked}}
\newcommand{\Certified}{\cert{Certified}}
\newcommand{\Assumed}{\cert{Assumed}}
\newcommand{\State}{\mode{State}}
\newcommand{\Observable}{\mode{Observable}}

\title[Code-Indexed Resource Calculus for FTQC]{A Code-Indexed Probabilistic Resource Calculus for Fault-Tolerant Quantum Programming}

\author{Anonymous Authors}
\affiliation{\institution{Anonymous Institution}\country{}}

\begin{document}

\begin{abstract}
Fault-tolerant quantum programs are written as logical operations, but they run through a heterogeneous collection of code-specific gates, code switches, magic-state factories, decoders, and layout constraints.
This mismatch makes a compiler decision hard to audit: a direct gate may be applied outside the current code's capability set, a probabilistic resource state may be reused as if it were classical data, or an imported physical theorem may be silently treated as a local proof.
This paper presents \system, a code-indexed, probabilistic, resource-aware calculus for elaborating logical quantum programs into certified fault-tolerant acquisition plans.
The calculus follows a simple running example: a \code{SurfaceD5} program that asks for $\gate{T}$, $\gate{CNOT}$, and $\gate{CCZ}$ operations even though its initial code exposes only a small direct gate set.
\system assigns encoded data types $Q[C,d,\mu]$, tracks magic and Bell resources linearly, and composes conservative effects for logical error, failure, acceptance, space, and time.
Its central judgment, $\judgment$, checks that each target acquisition step is justified by a capability rule, consumes resources exactly once, respects a fixed backend context, and records whether the rule is locally checked, certified by an imported theorem, or still assumed.
We implement the calculus in a prototype compiler/checker that plans direct gates, switch-via-code paths, resource-state injection, and hybrid regions over a small rule library drawn from recent fault-tolerant quantum protocols.
The implementation emits auditable sidecars, SMT obligations, protocol certificates, decoder contracts, geometry checks, and a Lean proof kernel for core resource and certificate properties.
On the current case studies, the checker accepts multiple exploratory acquisition strategies, rejects unsupported direct gates, duplicate resource consumption, and backend-incompatible qLDPC-style switches, and produces a full-stack report with 53 proof obligations and explicit imported-theorem boundaries.
\end{abstract}

\maketitle

\input{sections/introduction}
\input{sections/overview}
\input{sections/core-language}
\input{sections/typing-metatheory}
\input{sections/implementation}
\input{sections/case-study}
\input{sections/evaluation}
\input{sections/related-work}
\input{sections/limitations}
\input{sections/conclusion}

\begin{acks}
Omitted for anonymous review.
\end{acks}

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\appendix
\input{sections/appendix-claim-evidence}

\end{document}
